---
title: "Adding dynamic annotations to console outputs with the R package `annotated`"
author: "Maksim Rudnev"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding annotations to console outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


1. [Create annotations](#create-annotations)
2. [Getting your output annotated](#getting-your-output-annotated)
3. [Technical details](#technical-details)


## 1. Create annotations


One need to create an annotation or use the existing one. If you have an annotation files, check the next section - on how to use it.

Annotation is best created with a real examplary output. So create some call and put it unquoted to the `create_annotation()` function.


<pre><span style="color:blue">
library(annotated)
create_annotation(lm(dist ~ speed, cars))
</pre>

After this, you will be carried through an interactive Q&A asking you about each line of the output and how it should be annotated. If you opt for conditional annotations, the conditions should be also added.

**Annotations** are short fragments of text to help students (or yourself) to understand the output. They may contain fragment of R code enclosed in backticks, like in Rmarkdown.\* For example, if you can some additional information to the output, say CFI index: ``` by the way, CFI is `fitMeasures(ob)["cfi"]` ```. Try to avoid using `"` quotation marks (if it's unavoidable, escape it, i.e. - add a backslash before `\"` ).

Consider using package `crayon` to format annotations, for example: ```"No, you `crayon::bold("can not")` interpret these coefficients."```.

**Conditions** are optional pieces of R code that should return *TRUE* or *FALSE*, they may contain an outputted object referred to as `ob`, for example, `nobs(ob)<100` is a proper condition to show yout annotations if the model was fitted to sample less than 100. All conditions are enclosed in `if()` and executed before printing the corresponding annotation, if the expression returns *TRUE*. If one condition is specified than at least second should be specified as well, for example if there is a condition `nobs(ob)<100`, there should be also `nobs(ob)>100`.

<small>\* Scoping may still be a problem, so try to use in annotations only standard methods, or methods included in the same package as a function being annotated </small>

<details>
  <summary>
    Click here to see the full log of the interactive Q&A.
  </summary>

```

> create_annotation(lm(dist ~ speed, cars))
[1] 
[2] Call:
[3] lm(formula = dist ~ speed, data = cars)
[4] 
[5] Coefficients:
[6] (Intercept)        speed  
[7]     -17.579        3.932  
[8] 
Above is the output you are going to annotate.
Would you like to make annotations conditional?
You will be prompted to add conditions for every line of current output. 

1: Yes
2: No

Selection: 1
Every condition will be saved as an R script. You can refer to the output object of the initial call as 'ob'.
Now I am going to return it line by line so you can add your annotations.
If you do not want to comment the line, just press 'Enter'.
[1] 
[2] Call:
Is it conditional? 

1: Yes
2: No

Selection: 2
Annotate: 
[3] lm(formula = dist ~ speed, data = cars)
Is it conditional? 

1: Yes
2: No

Selection: 2
Annotate: This is a full call, including the arguments that weren't specified; those arguments were set to their default values.
[4] 
[5] Coefficients:
Is it conditional? 

1: Yes
2: No

Selection: 2
Annotate: These are unstandardized regression coefficients.
[6] (Intercept)        speed  
Is it conditional? 

1: Yes
2: No

Selection: 2
Annotate: 
[7]     -17.579        3.932  
Is it conditional? 

1: Yes
2: No

Selection: 1
What is condition 1? nobs(ob)<100
Annotate (if condition 1 is TRUE): The sample size is below 100 observations. The estimated parameters can be underpowered.
What is condition 2? nobs(ob)>99
Annotate (if condition 2 is TRUE): The sample size is 100 or higher.
Need to add another (last) condition? 

1: Yes
2: No

Selection: 2
[8] 
It's over. Press Enter to have a look at the annotated output. 
───────────────────────────────────────────────────────────────────────────────────────────────────────────────────── 
  Call: 
  lm(formula = dist ~ speed, data = cars) 
CONDITION: 
ANNOTATION: This is a full call, including the arguments that weren't sepcified; those arguments were set to their default values.

   
  Coefficients: 
CONDITION: 
ANNOTATION: These are unstandardized regression coefficients.

  (Intercept)        speed   
      -17.579        3.932   
CONDITION: nobs(ob)<100
ANNOTATION: The sample size is below 100 obersvations. `bold('The estimated parameters can be underpowered!')`

 CONDITION: nobs(ob)>99
ANNOTATION: The sample size is 100 or higher.

  
  ─────────────────────────────────────────────────────────────────────────────────────────────────────────────────────
How do I save the annotations?

1: + Save as R-script for future use with 'annotate' function.
2:   Save as html as the static example.

Enter one or more numbers separated by spaces, or an empty line to cancel
1: 1 2
To which functions this annotation is applicable? separate by spaces e.g., lm cfa:  lm
Give this file a name with .R extension: lm.annot.R

Entries are saved to lm.annot.R. You can manually edit them later.
Give this file a name with .html extension: example.html
all done.

```
</details>



After the tutorial finished, two files can be saved: **lm.annot.R** (see *Technical details* below) that can be used for dynamic annotation, and **example.html** that shows all annotations with conditions (as following):


<pre><br><span style='color:red;'></span>Call:<br><span style='color:red;'></span>lm(formula = dist ~ speed, data = cars)<br><span style='color:red;'>
ANNOTATION: This is a full call, including the arguments that weren't specified; those arguments were set to their default values.<br></span><br><span style='color:red;'></span>Coefficients:<br><span style='color:red;'>

ANNOTATION: These are unstandardized regression coefficients.<br></span>(Intercept)        speed  <br><span style='color:red;'></span>    -17.579        3.932  <br><span style='color:red;'>CONDITION1: nobs(ob)<100<br>ANNOTATION: The sample size is below 100 observations. `bold('The estimated parameters can be underpowered!')`<br>CONDITION2: nobs(ob)>99<br>ANNOTATION: The sample size is 100 or higher.<br></span><br><span style='color:red;'></span></pre>



## 2. Getting your output annotated

First, tell R where the libarary with annotations is located.

```{r eval = FALSE}

options(annotated.source="lm.annot.R")

```


Second, make a call to a function, in a usual way, but embraced by `annotated()`. Note that this would return the usual output of `lm()`, so you can save it to some object for further examination. For example, later you can check `summary(m)`

```{r eval = FALSE}

m <- annotated(lm(lifeExpF ~ ppgdp, UN))

```

<pre>
 <span style="color:blue"> > annotated(lm(lifeExpF ~ ppgdp, UN)) </span>

Call:
lm(formula = lifeExpF ~ ppgdp, data = UN)
 <span style="color:red">This is a full call, including the arguments that weren't specified; those arguments were set to their default values.</span>

Coefficients:
 <span style="color:red">These are unstandardized regression coefficients.</span>
(Intercept)        ppgdp  
  6.837e+01    3.018e-04  
 <span style="color:red"> The sample size is 100 or higher. </span>
</pre>


Likewise, if we fit the model to a sample less than 100, the annotations will change correspondingly: 

<pre>
 <span style="color:blue"> > annotated(lm(lifeExpF ~ ppgdp, UN[1:10,])) </span>

Call:
lm(formula = lifeExpF ~ ppgdp, data = UN)
 <span style="color:red">This is a full call, including the arguments that weren't specified; those arguments were set to their default values.</span>

Coefficients:
 <span style="color:red">These are unstandardized regression coefficients.</span>
(Intercept)        ppgdp  
  6.870e+01    3.374e-04   
 <span style="color:red">  The sample size is below 100 observations. <b>The estimated parameters can be underpowered!</b> </span>
</pre>


## 3. Technical details

`lm.annot.R` looks like this - it is a generated R code that creates a list with three elements: `annotation`, `conditions`, and `methods.for`. There are as many elements in each of `annotation` and `conditions` as long the the print output is. One can manually edit it.

```
#This is a script generated with an R package 'annotated'. It is supposed to be used as a library of annotations within the function 'annotated()'. Created on: 2019-02-10 20:48:35

list(
 annotation = list(
  cond1 = c(
    "",
    "",
    "This is a full call, including the arguments that weren't specified; those arguments were set to their default values.",
    "",
    "These are unstandardized regression coefficients.",
    "",
    "The sample size is below 100 observations. `red('The estimated parameters can be underpowered!')`",
    ""),
  cond2 = c(
    "",
    "",
    "",
    "",
    "",
    "",
    "The sample size is 100 or higher.",
    ""),
  cond3 = c(
    "",
    "",
    "",
    "",
    "",
    "",
    "",
    "")
 ),
 conditions = list(
  cond1 = c(
    "",
    "",
    "",
    "",
    "",
    "",
    "nobs(ob)<100",
    ""),
  cond2 = c(
    "",
    "",
    "",
    "",
    "",
    "",
    "nobs(ob)>99",
    ""),
  cond3 = c(
    "",
    "",
    "",
    "",
    "",
    "",
    "",
    "")
 ),
methods.for = "lm"
)
```


